Towards Comparable Active Learning

### 2023-02-23 doc/al_benchmark_paper.tex
docs

\documentclass[]{article}
\usepackage[a4paper, total={6.5in, 8.5in}]{geometry}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{float}
\usepackage{algorithm}
\usepackage{algorithmic}

@@

\subsection{Problem Description / Delineation}
\textbf{Basic Classification}\\
We assume a dataset $\mathcal{D} := (\mathcal{U}, \mathcal{L})$ consisting of a large pool of unlabeled data $\mathcal{U}$ and a small set of labeled data called the seed set with $|\mathcal{L}| \ll |\mathcal{U|}$. 
For the purpose of this work, only fully labeled datasets are used and the labels for $\mathcal{U}$ are suppressed until they are selected for labeling.
For evaluation purposes we also assume a held-out test set $(x_{test}, y_{test}) \in (\mathbb{R}^M, \mathbb{R}^C)$
We consider only classification problems, hence the instances of a dataset have the form $x \in \mathbb{R}^M$ for $\mathcal{U}$ and $(x, y) \in (\mathbb{R}^M, \mathbb{R}^C)$ for $\mathcal{L}$.
To perform classification, a model $\hat y_\theta := \mathbb{R}^M \rightarrow \mathbb{R}^C$ is used. To fit the model, it is parameterized by $\theta$ and subjected to loss $l(y, \hat y_\theta(x)) := \mathbb{R}^C \times \mathbb{R}^C \rightarrow \mathbb{R}$. For this work, categorical cross-entropy (CE) is used.
For evaluating classification performance, we use Accuracy \\ [1mm]
%
\textbf{Pool-based AL with single instances (non-batch setting)}\\
Active learning is defined as sequentially removing single instances $x_i \in \mathcal{U}_i\hspace{1mm}; \hspace{2mm} \mathcal{U}_{i+1} := \mathcal{U}_i / \{x_i\}$, requesting their label $y_i$ and adding them to the labeled pool $\mathcal{L}_{i+1} := \mathcal{L}_i \cup (x_i, y_i)$ until a fixed budget $B$ is exhausted $i := 0 \ldots B$.
After each added instance the classification model is retrained according to section \ref{sec:training_the_classifier} and its performance is measured on a held-out test set.
The quality of an active learning algorithm is evaluated by an "anytime" protocol that incorporates classification performance at every iteration, not just the final performance after the budget is exhausted.
We employ the normalized area under the accuracy curve (AUC):
\begin{equation}
	\operatorname{auc}(\mathcal{U}, \mathcal{L}, \hat y, B) := \frac{1}{B} \sum_{i=1}^{B} acc(y_{test}, \hat y_i(x_{test}))
\end{equation}
Where $\hat y_i$ is the retrained classification model after the i-\textit{th} instance was selected. \\ [1mm]
%
\textbf{Framing AL as RL}\\
We define the active learning process as an adapted reinforcement learning loop $(S, A, \tau, \Omega, \omega)$ where an environment iteratively will expose a state $s \in S$ to an agent $\Omega$, which will choose actions $a \in A$.
For each iteration $i$ the environment samples a subset of size $\tau$ of unlabeled instances $x_i \underset{\tau}{\sim} \mathcal{U}_i$, constructs the state $s_i := \omega(x_i)$ and presents it to the agent to select an action $a_i := \Omega(s_i)$.
The action $a_i$ is the index of the selected instance in $u_i$ out of all possible indices $A := [0 \ldots \tau]$.
This process is repeated $B$ times $i := [0 \ldots B]$.

@@

>>>>

\documentclass[]{article}
\usepackage[a4paper, total={6.5in, 8.5in}]{geometry}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{float}
\usepackage{algorithm}
%\usepackage{algorithmic}
\usepackage{algpseudocode}

@@

\subsection{Problem Description / Delineation}
\textbf{Basic Classification}\\
We assume a dataset $\mathcal{D} := (x_i, y_i) \hspace{0.5mm}; \hspace{1mm} i := 0 \ldots N$ consisting of instances $x_i \in \mathbb{R}^M$ and corresponding $y_i \in \mathbb{R}^C$.
For evaluation purposes we assume a held-out test set $\mathcal{D}^{\text{test}}$ with the same characteristics.
We consider classification problems with one-hot encoded classes, hence $C$ models the number of classes.
To perform classification, a model $\hat y_\theta : \mathbb{R}^M \rightarrow \mathbb{R}^C$ is used. To fit the model, it is parameterized by $\theta$ and subjected to loss $\ell: \mathbb{R}^C \times \mathbb{R}^C \rightarrow \mathbb{R}$. For this work, categorical cross-entropy (CE) is used.
For evaluating classification performance, we use accuracy on the test set $\operatorname{Acc}(\mathcal{D}^{\text{test}}, \hat{y}_\theta)$. \\ [1mm]
%
\textbf{Pool-based AL with single instances (non-batch setting)}\\
To construct the active learning setting, we suppress the labels $y_i$ of $\mathcal{D}$ to form the unlabeled pool $\mathcal{U} := x_i \hspace{0.5mm}; \hspace{1mm} i := 0 \ldots N$ and form and initial labeled pool $\mathcal{L}$ by uniformly sampling $k$ number of instances per class from $\mathcal{U}$ and recovering their label. 
The result of this so-called "seeding" process is $\mathcal{L} := (x_i, y_i) \hspace{0.5mm}; \hspace{1mm} i := 0 \ldots k*C$. \\
Active learning is defined as sequentially removing single instances $x_i \in \mathcal{U}_i\hspace{1mm}; \hspace{2mm} \mathcal{U}_{i+1} := \mathcal{U}_i \setminus´ \{x_i\}$, recovering their label $y_i$ and adding them to the labeled pool $\mathcal{L}_{i+1} := \mathcal{L}_i \cup (x_i, y_i)$ until a fixed budget $B$ is exhausted $i := 0 \ldots B$.
After each added instance the classification model is retrained according to section \ref{sec:training_the_classifier} and its performance is measured on a held-out test set.
The quality of an active learning algorithm is evaluated by an "anytime" protocol that incorporates classification performance at every iteration, not just the final performance after the budget is exhausted.
We employ the normalized area under the accuracy curve (AUC):
\begin{equation}
	\operatorname{auc}(\mathcal{U}, \mathcal{L}, \hat y, B) := \frac{1}{B} \sum_{i=1}^{B} \operatorname{Acc}(y_{test}, \hat y_i(x_{test}))
\end{equation}
Where $\hat y_i$ is the retrained classification model after the i-\textit{th} instance was selected. \\ [1mm]
%
\textbf{Framing AL as RL}\\
We define the active learning process as an adapted reinforcement learning loop $(S, A, \tau, \Omega, \omega)$ where an environment iteratively will expose a state $s \in S$ to an agent $\Omega$, which will choose actions $a \in A$.
For each iteration $i$ the environment samples a subset of size $\tau$ of unlabeled instances $x^{(i)} \underset{\tau}{\sim} \mathcal{U}^{(i)}$, constructs the state $s^{(i)} := \omega(x^{(i)})$ and presents it to the agent to select an action $a^{(i)} := \Omega(s^{(i)})$.
The action $a^{(i)}$ is the index of the selected instance in $u^{(i)}$ out of all possible indices $A := [1 \ldots \tau]$.
This process is repeated $B$ times $i := [1 \ldots B]$.

\begin{minipage}{0.59\linewidth}
	\begin{algorithm}[H]
		\caption{Active Learning}\label{alg:active_learning}
		\begin{algorithmic}[1]
			\Require $\mathcal{U}$ \Comment{Unlabeled Pool}
			\Require $\tau$ \Comment{Unlabeled Sample Size}
			\Require $\Omega$ \Comment{AL Agent}
			\Require $\omega$ \Comment{Environment State function}
			\State $\mathcal{L}^{(1)} \gets \operatorname{seed}(\mathcal{U})$  \Comment{Create the initial labeled set}
			\State $\mathcal{U}^{(1)} \gets \mathcal{U}$
			\For{$i := 1 \ldots B$}
			\State $u^{(i)} \underset{\tau}{\sim} \mathcal{U}^{(i)}$
			\State $s^{(i)} \gets \omega(u^{(i)})$
			\State $a^{(i)} \gets \Omega(s^{(i)})$ \Comment{$a^{(i)}$ is an index inside of $u^{(i)}$}
			\State $y^{(i)} \gets \operatorname{oracle}(u^{(i)}_{a})$ \Comment{$u^{(i)}_{a}$ is shorthand for $u^{(i)}_{a^{(i)}}$}
			\State $\mathcal{L}^{(i+1)} \gets \mathcal{L}^{(i)} \cup \{(u^{(i)}_a, y^{(i)})\}$
			\State $\mathcal{U}^{(i+1)} \gets \mathcal{U}^{(i)} \setminus \{u^{(i)}_a\}$
			\State $\text{acc}^{(i)} \gets \operatorname{Retrain}(\mathcal{L}^{(i+1)})$  \Comment{$\operatorname{Retrain}(\mathcal{L}^{(i+1)})$ is shorthand for $\operatorname{Retrain}(\mathcal{L}^{(i+1)}, \mathcal{L}^\text{test}, \hat y_\theta, e^\text{max})$}
			\EndFor
			\State
			\Return $\frac{1}{B} \sum_{i=1}^{B} \text{acc}^{(i)}$
		\end{algorithmic}
	\end{algorithm}
\end{minipage}
\hspace{3mm}
\begin{minipage}{0.35\linewidth}
	\begin{algorithm}[H]
		\caption{Retrain}\label{alg:retrain}
		\begin{algorithmic}[1]
			\Require $\mathcal{L}$ \Comment{Labeled Pool}
			\Require $\mathcal{L}^\text{test}$ \Comment{Labeled Test Data}
			\Require $\hat y_\theta$ \Comment{Classification Model}
			\Require $e^\text{max}$ \Comment{Maximum Epochs}
			\State $\text{loss}^* \gets \infty$
			\For{$i := 1 \ldots e^{\text{max}}$}
			\State $\theta_{i+1} \gets \theta_i - \eta \nabla_\theta \ell(\mathcal{L}, \hat y_{\theta})$
			\State $\text{loss}_i \gets \ell(\mathcal{L}^\text{test}, \hat y_{\theta})$
			\If{$\text{loss}_i < \text{loss}^*$}
			\State $\text{loss}^* \gets \text{loss}_i$
			\Else
			\State Break
			\EndIf
			\EndFor
			\State
			\Return Acc($\mathcal{L}^\text{test}, \hat y_{\theta}$)
		\end{algorithmic}
	\end{algorithm}
\end{minipage}

@@

\appendix

\section{Comparison of different sample sizes $\tau$}
\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\linewidth]{img/tau_ablation.png}
\end{figure}
